Approximation as Differentiation

### 2025-07-09 main/related-work.tex
WIP.
fam -> indexed-family.

Galois slicing was introduced by \citet{perera12a} as an operational approach to program slicing for pure
functional programs, based on Galois connections between lattices of input and output approximations. A
connection to stable functions in relation to minimal slices for short-circuiting operations was alluded to
in~\citet{perera13}. Subsequent work extended the approach to languages with assignment and
exceptions~\cite{ricciotti17} and concurrent systems, applying Galois slicing to the
$\pi$-calculus~\cite{perera16d}. For the $\pi$-calculus the analysis shifted from functions to transition
relations, considering individual transitions $P \longrightarrow Q$ between configurations $P$ and $Q$ as
analogous to the edge between $x$ and $f(x)$ in the graph of $f$, and building Galois connections between
$\downset{P}$ and $\downset{Q}$. The main difference with the approach presented here is that the earlier work
also computes \emph{program slices}, using approximation lattices that represent partially erased programs; we
discuss this further in \secref{conclusion} below.

@@

\emph{Tangent categories}, due originally to \citet{rosický84} and developed by \citet{cockett14,cockett18},
provide an abstract categorical framework for reasoning about differentiation, inspired by the structure of
the tangent bundle in differential geometry. In a tangent category, each object $X$ is equipped with a tangent
bundle $T(X)$, and each morphism $f: X \to Y$ has a corresponding differential map $T(f): TX \to TY$
satisfying axioms analogous to the chain rule and linearity of differentiation. Tangent categories generalise
Cartesian differential categories \cite{cdcs}, which model differentiation over cartesian closed categories
using a syntactic derivative operator. The motivation behind tangent categories is to \roly{Bob to say
something here..}

>>>>

Galois slicing was introduced by \citet{perera12a} as an operational approach to program slicing for pure
functional programs, based on Galois connections between lattices of input and output approximations. A
connection to stable functions in relation to minimal slices for short-circuiting operations was alluded to
in~\citet{perera13}, but not explored. Subsequent work extended the approach to languages with assignment and
exceptions~\cite{ricciotti17} and concurrent systems, applying Galois slicing to the
$\pi$-calculus~\cite{perera16d}. For the $\pi$-calculus the analysis shifted from functions to transition
relations, considering individual transitions $P \longrightarrow Q$ between configurations $P$ and $Q$ as
analogous to the edge between $x$ and $f(x)$ in the graph of $f$, and building Galois connections between
$\downset{P}$ and $\downset{Q}$. The main difference with the approach presented here is that the earlier work
also computes \emph{program slices}, using approximation lattices that represent partially erased programs; we
discuss this further in \secref{conclusion} below.

@@

\emph{Tangent categories}, due originally to \citet{rosický84} and developed by \citet{cockett14,cockett18},
provide an abstract categorical framework for reasoning about differentiation, inspired by the structure of
the tangent bundle in differential geometry. In a tangent category, each object $X$ is equipped with a tangent
bundle $T(X)$, and each morphism $f: X \to Y$ has a corresponding differential map $T(f): T(X) \to T(Y)$
satisfying axioms analogous to the chain rule and linearity of differentiation. Tangent categories generalise
Cartesian differential categories \cite{cdcs}, which model differentiation over cartesian closed categories
using a syntactic derivative operator. The motivation behind tangent categories is to \roly{Bob to say
something here..}
